### **Lead in WA Schools: Making Figures**
This file is used to make plots from a cleaned dataset that includes data from DOH lead water testing, OSPI student enrollment/demographic information, leaded flights, blood lead levels, and lead risk from housing. To run this file, you will need to place the following files in your working directory:
(1) "lead_and_ospi.csv"

### **Step 1. Load R Libraries and Upload Cleaned Data Resulting from "Clean_Lead_Data.Rmd" file**
```{r}
library(tidyr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(sf)
library(scales)

setwd(dir = "~/Dropbox/Katya Dropbox/HCDE pre-tenure (2023-2028)/Research/Lead in WA Schools/")

lead_and_ospi <-read.csv(file = "lead_and_ospi.csv", sep = ",",stringsAsFactors = T)
sd_map <- read_sf("datasets/mapping_geospatial_WA/Geojson Files/Washington_School_District_Boundaries.geojson")

```

### **Step 2. Calculate % Lead Contamination Above Thresholds at the level of school districts, schools, and fixtures**
```{r}
district_contamination <- lead_and_ospi %>%
  group_by(school_district) %>%
  summarise(total_samples = n(),
            "1 ppb" = 100 * sum(mean_lead_result_ppb > 1, na.rm = TRUE)/total_samples,
            "5 ppb" = 100 * sum(mean_lead_result_ppb > 5, na.rm = TRUE)/total_samples,
            "10 ppb" = 100 * sum(mean_lead_result_ppb > 10, na.rm = TRUE)/total_samples)

district_contamination <- district_contamination %>%
  pivot_longer(cols=c(3:5),
               names_to = "Lead Contamination Threshold",
               values_to = "Percent Contamination (%)")  

school_contamination <- lead_and_ospi %>%
  filter(school_name != "Demolished School") %>%
  group_by(school_district, 
           school_name,
           DOH_testing_round,
           x,
           y) %>%
 summarise(total_samples = n(),
            "1 ppb" = 100 * sum(mean_lead_result_ppb > 1, na.rm = TRUE)/total_samples,
            "5 ppb" = 100 * sum(mean_lead_result_ppb > 5, na.rm = TRUE)/total_samples,
            "10 ppb" = 100 * sum(mean_lead_result_ppb > 10, na.rm = TRUE)/total_samples)

school_contamination <- school_contamination %>%
  pivot_longer(cols=c(7:9),
            names_to = "Lead Contamination Threshold",
               values_to = "Percent Contamination (%)")  
               
fixture_contamination <- lead_and_ospi %>%
filter(school_name != "Demolished School") %>%
group_by(fixture_type) %>%
summarise(total_samples = n(),
            "1 ppb" = 100 * sum(mean_lead_result_ppb > 1, na.rm = TRUE)/total_samples,
            "5 ppb" = 100 * sum(mean_lead_result_ppb > 5, na.rm = TRUE)/total_samples,
            "10 ppb" = 100 * sum(mean_lead_result_ppb > 10, na.rm = TRUE)/total_samples)

fixture_contamination <- fixture_contamination %>%
  pivot_longer(cols=c(3:5),
         names_to = "Lead Contamination Threshold",
               values_to = "Percent Contamination (%)")  
```