---
title: "trisha-final"
output: html_document
date: "2025-08-15"
---

###**Setup R Markdown file, set working directory**
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/Desktop/Lead in WA Schools")
knitr::opts_chunk$set(echo = TRUE)

```

###**Add libraries and data files**
```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(sf)
library(patchwork)

sd_map <- read_sf("datasets/mapping_geospatial_WA/Geojson Files/Washington_School_District_Boundaries.geojson")
esd_data <- read_sf("datasets/mapping_geospatial_WA/Geojson Files/Washington_Educational_Service_District_Boundaries.geojson")
county_data <- read_sf("datasets/mapping_geospatial_WA/Geojson Files/WA_County_Boundaries.geojson")

all_lead_datasets <- read.csv(file = "all_lead_datasets.csv", sep = ",",stringsAsFactors = T)

```



###**Create general data frames (lucy)**
```{r}

all_lead_datasets <- all_lead_datasets %>% 
  mutate(school_county = if_else(school_county == "PendOreille",
                                 "Pend Oreille",
                                 school_county))

county_boundaries <- county_data %>% 
  rename(school_county = JURISDICT_LABEL_NM)

bll_data <- all_lead_datasets %>% 
  select(school_county,
         num_children_tested,
         num_with_bl_results_above_5,
         num_children_below_72_months,
         percent_with_bl_result_above_5,
         reliability) %>% 
  filter(!is.na(num_children_below_72_months)) %>%
  distinct() %>% 
  left_join(county_boundaries, by = "school_county") %>% 
  st_as_sf(sf_column_name = "geometry", crs = 4326)

lead_water_data <- all_lead_datasets %>% 
  select(x,
         y,
         school_name,
         DOH_testing_round,
         school_district,
         school_county,
         building_id,
         fixture_type,
         fixture_housing_location,
         number_replicates,
         mean_lead_result_ppb,
         contamination_status,
         building_construction_year)%>% 
  left_join(county_boundaries, by = "school_county") %>% 
  st_as_sf(sf_column_name = "geometry", crs = 4326)

housing_risk_data <- all_lead_datasets %>% 
  select(school_county,
         num_units_w_lead_hazard_estimated,
         num_total_housing_units,
         pct_units_w_lead_hazard_estimated) %>% 
  filter(!is.na(num_units_w_lead_hazard_estimated)) %>%
  distinct() %>% 
  left_join(county_boundaries, by = "school_county") %>%
  st_as_sf(sf_column_name = "geometry", crs = 4326)

median_income_data <- all_lead_datasets %>% 
  group_by(school_county) %>% 
  summarise(median_hh_income_2022 = first(na.omit(median_hh_income_2022))) %>% 
  left_join(county_boundaries, by = "school_county")

flights_data <- all_lead_datasets %>% 
  select(x,
         y,
         school_name,
         school_district,
         school_county,
         total_leaded_flights) %>% 
  filter(school_county == "King") %>% 
  distinct()

```


###**BLL and Lead Fixtures: Makes map of WA state with county boundaries. Color of county represents the percent of children with high blood lead levels and the size of the red dot represents the average percent of fixtures with elevated lead levels across the county. (lucy)**

```{r}

# Note: creating county_ave_lead turns ~30000 rows into 37, it takes about 20 minutes on my computer -lucy
county_ave_lead <- lead_water_data %>%
  group_by(school_county) %>%
  summarise(avg_pct_above_5 = mean(mean_lead_result_ppb > 5, na.rm = TRUE) * 100,
            avg_lead_level = mean(mean_lead_result_ppb),
            .groups = "drop")

county_ave_lead_centroids <- county_ave_lead %>% 
  st_centroid()

map_both <- ggplot() +
  geom_sf(data = bll_data, aes(fill = percent_with_bl_result_above_5)) +
  geom_sf(data = county_ave_lead_centroids, aes(size = avg_pct_above_5), color = "red", alpha = 0.7) +
  scale_fill_viridis_c(option = "viridis", na.value = "gray80", name = "% of children < 72 months with BLL >= 5ug/Dl") +  
  scale_size_continuous(name = "Avg % fixtures >= 5ppb") +
  labs(title = "Blood Lead Levels in WA by County") +
  theme_bw()

map_both

ggsave("Trisha_Lucy_Figures/county_bl_levels_with_fixtures.png", plot = map_both, bg = "white", width = 8, height = 6)

```

###**BLL by County: Makes a map of WA state with county boundaries. Color of each county represents the percent of children with high blood lead levels.(lucy)**
```{r}
#This map is missing two counties that aren't accounted for in the school testing data. San Juan county and Wahkiakum county. 

map_bll_county <- ggplot() +
  geom_sf(data = bll_data, aes(fill = percent_with_bl_result_above_5), col = "white") +
  scale_fill_viridis_c(option = "viridis", na.value = "gray80", name = "% of children < 72 months with BLL >= 5ug/Dl") +  
  labs(title = "Blood Lead Levels in WA by County") +
  theme_bw()

map_bll_county

ggsave("Trisha_Lucy_Figures/county_bll.png", plot = map_bll_county, bg = "white", width = 8, height = 6)

```

###**Lead in Fixtures by County: Makes a map of WA state with county boundaries. Color of each county represents the average percent of fixtures with lead levels greater than 5 ppb.(lucy)**
```{r}
#This map is missing two counties that aren't accounted for in the school testing data. San Juan county and Wahkiakum county. 

map_lead_fixtures_county <- ggplot() + 
  geom_sf(data = county_ave_lead, aes(fill = avg_pct_above_5), col = "white") +
  scale_fill_viridis_c(option = "viridis", na.value = "gray80", name = "avg % of fixtures > 5 ppb") +
  labs(title = "% of fixtures with more than 5 ppb lead by county") +
  theme_bw()

map_lead_fixtures_county

ggsave("Trisha_Lucy_Figures/map_pct_fixtures_more_than_5_county.png", plot = map_lead_fixtures_county, width = 8, height = 6)

```

###**Combined: Combines two maps vertically: 1) map_bll_county — percent of children with high blood lead levels by county and 2) map_lead_fixtures_county — average percent of fixtures with >5 ppb lead by county.(lucy)**
```{r}
combined <- map_bll_county / map_lead_fixtures_county

combined

ggsave("Trisha_Lucy_Figures/two_graph_bll_fixtures_by_county.png", plot = combined, width = 8, height = 6)

```

###**Scatterplot: Creates a scatter plot showing the relationship between: 1) Average % of fixtures with lead levels > 5 ppb (x-axis) and 2) Average % of children <72 months with blood lead levels ≥ 5 µg/dL (y-axis). Each point represents one county. (lucy)**
```{r}
fixtures_no_location <- county_ave_lead %>% 
  select(school_county,
         avg_pct_above_5,
         avg_lead_level) %>% 
  st_drop_geometry()

bll_no_location <- bll_data %>% 
  select(school_county,
         percent_with_bl_result_above_5) %>% 
  st_drop_geometry()

bll_fixtures_combined <- bll_no_location %>% 
  left_join(fixtures_no_location, by = "school_county")

scatter <- bll_fixtures_combined %>% 
  ggplot(aes(x = avg_pct_above_5, y = percent_with_bl_result_above_5)) +
  geom_point(color = "blue", size = 3) +
  labs(title = "Average % of fixtures >5pbb vs Avg % children ≥ 5ug/Dl by County",
       x = "Average % of fixtures > 5 ppb", 
       y = "Average % children <72 months with >= 5ug/Dl BLL") +
  theme_bw()

scatter

ggsave("Trisha_Lucy_Figures/scatter_bll_vs_fixtures_county.png", plot = scatter, width = 8, height = 6)

```

###**Housing County Map:Maps housing lead hazard risk data by county. (lucy)**
```{r}
# These maps are missing two counties that aren't accounted for in the school testing data. San Juan county and Wahkiakum county. 

# map % units with lead hazard
map_housing_risk <- ggplot() + 
  geom_sf(data = housing_risk_data, aes(fill = pct_units_w_lead_hazard_estimated), col = "white") +
  scale_fill_viridis_c(option = "viridis", na.value = "gray80", name = "avg % units with lead hazard estimated") +
  labs(title = "% units with lead hazard per county") +
  theme_bw()

map_housing_risk

# combines BLL map and housing risk map
bll_housing_combined <- map_bll_county / map_housing_risk

bll_housing_combined

# saves single and combined map
ggsave("Trisha_Lucy_Figures/housing_risk_by_county.png", plot = map_housing_risk, width = 8, height = 6)
ggsave("Trisha_Lucy_Figures/bll_housing.png", plot = bll_housing_combined, width = 8, height = 6)

```

###**Scatter plot of building construction year vs lead concentration (2022): Creates a scatter plot showing lead concentration (ppb) by building constructionyear for each sample in 2022. Points are colored by fixture type. Plot 1 includes all fixtures and a red horizontal line at 5 ppb.Plot 2 includes all fixtures with lead levels under 100 ppb and a red horizontal line at 5 ppb. Plot 3 includes all fixtures with lead leves under 100 ppb and deoes not include any red horizontal lines. (lucy)**
```{r}
#creates data frame
lead_and_building_2022 <- all_lead_datasets %>% 
  filter(DOH_testing_round == 2022,
         building_construction_year != 1111,
         ) %>% 
  select(x,
         y,
         school_name,
         school_district,
         school_county,
         fixture_type,
         fixture_housing_location,
         fixture_housing_type,
         number_replicates,
         mean_lead_result_ppb,
         contamination_status,
         building_detail,
         building_construction_year)

#plot of all samples with horizontal line
by_sample_all <- lead_and_building_2022 %>% 
  ggplot() +
  geom_point(aes(x = building_construction_year, y = mean_lead_result_ppb, color = fixture_type)) +
  geom_hline(yintercept = 5, color = "red", linewidth = 1) + 
  labs(title = "Building Year vs. Lead Concentration by sample in 2022 \n", x = "\n Building Construction Year",
       y = "Lead Concentration (ppb) \n") +
  scale_color_manual(values = c(
    "Bottle Refill Station" = "#0072B2",
    "Tap/Sink" = "#E69F00",
    "Ice Machine/Fridge" = "#009E73",
    "Pot Filler" = "#F0E442",
    "Other" = "#CC79A7",
    "Water Fountain" = "#56B4E9",
    "Water Cooler" = "#D55E00",
    "Sprayer/Hose" = "#000000"
  )) +
  theme_bw()

by_sample_all

ggsave("Trisha_Lucy_Figures/building_year_vs_lead_all.png", plot = by_sample_all, width = 8, height = 6)

#plot of all samples under 100 ppb with horizontal line
lead_and_building_2022_filtered <- lead_and_building_2022 %>%
  filter(mean_lead_result_ppb < 100)

by_sample_under_100 <- lead_and_building_2022_filtered %>% 
  ggplot() +
  geom_point(aes(x = building_construction_year, y = mean_lead_result_ppb, color = fixture_type)) +
  geom_hline(yintercept = 5, color = "red", linewidth = 1) + 
  labs(title = "Building Year vs. Lead Concentration by sample in 2022 \n", x = "\n Building Construction Year",
       y = "Lead Concentration (ppb) \n") +
  scale_color_manual(values = c(
    "Bottle Refill Station" = "#0072B2",
    "Tap/Sink" = "#E69F00",
    "Ice Machine/Fridge" = "#009E73",
    "Pot Filler" = "#F0E442",
    "Other" = "#CC79A7",
    "Water Fountain" = "#56B4E9",
    "Water Cooler" = "#D55E00",
    "Sprayer/Hose" = "#000000"
  )) +
  theme_bw()

by_sample_under_100

ggsave("Trisha_Lucy_Figures/building_year_vs_lead_under_100.png", plot = by_sample_under_100, width = 8, height = 6)

#plot of all fixtures under 100 ppb without horizontal line
by_sample_under_100_no_bar <- lead_and_building_2022_filtered %>% 
  ggplot() +
  geom_point(aes(x = building_construction_year, y = mean_lead_result_ppb, color = fixture_type)) +
  labs(title = "Building Year vs. Lead Concentration by sample in 2022 \n", x = "\n Building Construction Year",
       y = "Lead Concentration (ppb) \n") +
  scale_color_manual(values = c(
    "Bottle Refill Station" = "#0072B2",
    "Tap/Sink" = "#E69F00",
    "Ice Machine/Fridge" = "#009E73",
    "Pot Filler" = "#F0E442",
    "Other" = "#CC79A7",
    "Water Fountain" = "#56B4E9",
    "Water Cooler" = "#D55E00",
    "Sprayer/Hose" = "#000000"  
    )) +
  theme_bw()

by_sample_under_100_no_bar

ggsave("Trisha_Lucy_Figures/building_year_vs_lead_under_100_no_bar.png", plot = by_sample_under_100_no_bar, width = 8, height = 6)

```

###**Plotting a geospatial map of WA state with markers of which districts have been tested in which years, and markers for every school tested, with size of the marker indicating the number of samples per school.(trisha)**
```{r}


#standardizing names of districts
sd_map <- sd_map %>%
  mutate(district_name = ShortName)

#create tested flags
tested_22 <- all_lead_datasets %>%
  filter(DOH_testing_round == "2022") %>%
  distinct(school_district)
tested_18 <- all_lead_datasets %>%
  filter(DOH_testing_round == "2018") %>%
  distinct(school_district)

sd_map <- sd_map %>%
  mutate(
    tested_18 = district_name %in% tested_18,
    tested_22 = district_name %in% tested_22,
    tested_year = case_when(
      tested_18 & tested_22 ~ "Both",
      tested_22 ~ "2021",
      tested_18 ~ "2018",
      TRUE ~ "None"
    )
  )

# Filter school coordinates within WA bounds and add year tag
schools_data <- all_lead_datasets
schools_data <- schools_data%>%
  drop_na(x, y)%>%
  filter(
    #filtering in a bounding box for washington state to remove outliers
    x >= -125 & x <= -116,
    y  >=  45.5 & y  <=  49.1
  )

# Convert schools to sf
# Count number of samples per school
school_sample_counts <- schools_data %>%
  group_by(school_name) %>%
  summarise(n_samples = n(), .groups = "drop")

# Join back to main data
schools_data <- schools_data %>%
  left_join(school_sample_counts, by = "school_name")

schools_sf <- st_as_sf(schools_data, coords = c("x", "y"), crs = 4326)


# --- Add centroids for labeling districts safely ---

# Project to WA State Plane CRS for accurate centroid calculation
WSD_boundaries_proj <- st_transform((sd_map), 2855)

# Compute centroids in projected space
WSD_boundaries_proj <- WSD_boundaries_proj %>%
  mutate(label_geometry = st_centroid(geometry))

# Transform back to EPSG:4326 for plotting
WSD_boundaries <- st_transform(WSD_boundaries_proj, 4326)

# Plot
ggplot() + 
  geom_sf(
    data = sd_map
  ) +
  geom_sf(
    data = schools_sf,
    aes(size = n_samples),
    shape = 21,
    stroke = 0.4
  ) + 
  scale_fill_manual(
      values = c(
        "None" = "gray90",
        "2018" = "#78AA83",
        "2022" = "#F2ADAD",
        "Both" = "#8B7AAE"
      ),
      name = "DOH School Water \nLead Sampling", 
      labels = c(
        "None" = "Not tested", 
        "2018" = "Tested in 2018", 
        "2022" = "Tested in 2022",
        "Both" = "Tested in 2018 & 2022"
      ) 
  ) +
  theme_void() + 
  theme(
    legend.position = "none",  # <- This line removes the legend
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(title = "WA Schools Water Testing\n", size = NULL)
```

###**Making a facet wrapped graph comparing lead hazards to median income (trisha)**
```{r}

#establishing variables from not-common datasets
median_income <- all_lead_datasets %>%
  select(school_county, median_hh_income_2022)

lead_contamination <- #combined_lead_data %>%
  select(school_county, pct_above_5)%>%
  group_by(school_county) %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE))%>%
  drop_na()

housing_risk <- all_lead_datasets%>%
  select(school_county, pct_units_w_lead_hazard_estimated)

#making a combined dataframe
combined <- housing_risk %>%
  inner_join(median_income, by="county")%>%
  inner_join(lead_contamination, by="county")


plot_data <- bind_rows(
  housing_risk %>% mutate(metric = "% Lead Risk in Housing"),
  bll_levels %>% mutate(metric = "% BLL >5µg/dL in Kids <6"),
  lead_contamination %>% mutate(metric = "% Lead Contamination in Schools")
)

# Split the data into 3 subsets
lead_plot <- plot_data %>% filter(metric == "% Lead Risk in Housing")
bll_plot <- plot_data %>% filter(metric == "% BLL >5µg/dL in Kids <6")
school_plot <- plot_data %>% filter(metric == "% Lead Contamination in Schools")

# Plot 1
p1 <- ggplot(lead_plot, aes(x = med_income, y = value, color = med_income)) +
  #finds the R^2 value for the linear regression
  stat_poly_eq(
    aes(group = 1, label = paste("bold(", ..rr.label.., ")")),
    formula = y ~ x,
    parse = TRUE,
    label.x = 110000,
    label.y =9,
    size = 3,
    color = "gray30",
    geom = "label",
    color = "black",        # text color
    fill = "white",         # background box color
    label.size = 0,
    label.padding = unit(0.15, "lines")  # tighter box
  ) +
  geom_text_repel(
    aes(label = county),
    size = 2.5,
    color = "black",
    box.padding = 0.3,
    nudge_y = 0.1,
    segment.size = 0 
  ) +
  geom_smooth(method = "lm", se = FALSE, color = scales::alpha("gray30", 0.6),, linewidth = 0.5) +
  geom_point(size = 2) +
  scale_color_viridis_c(option = "C", direction = -1, name = "Median Income") +
  scale_x_continuous(
    labels = scales::dollar_format(scale = 0.0001, suffix = "0K"),
    breaks = seq(60000, 130000, 10000)
  ) +
  labs(x = NULL, y = "Lead Exposure Risk from \npre-1978 Housing (%)") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(),
    panel.grid.major = element_blank(),  # removes major grid lines
    panel.grid.minor = element_blank(),  # already in your code; keeps it
    axis.title.y = element_text(margin = margin(r = 10))
  )

# Plot 2
p2 <- ggplot(bll_plot, aes(x = med_income, y = value, color = med_income)) +
  stat_poly_eq(
    aes(group = 1, label = paste("bold(", ..rr.label.., ")")),
    formula = y ~ x,
    parse = TRUE,
    label.x = 110000,
    label.y = 0.5,
    size = 3,
    color = "gray30",
    geom = "label",
    color = "black",        # text color
    fill = "white",         # background box color
    label.size = 0,
    label.padding = unit(0.15, "lines")  # tighter box
  ) +
  geom_text_repel(
    aes(label = county),
    size = 2.5,
    color = "black",
    box.padding = 0.3,
    nudge_y = 0.3,
    segment.size = 0 
  ) +
  geom_smooth(method = "lm", se = FALSE, color = scales::alpha("gray30", 0.6), linewidth = 0.5) +
  geom_point(size = 2) +
  scale_color_viridis_c(option = "C", direction = -1, name = "Median Income") +
  scale_x_continuous(
    labels = scales::dollar_format(scale = 0.0001, suffix = "0K"),
    breaks = seq(60000, 130000, 10000)
  ) +
  labs(x = NULL, y = "Blood Lead Levels in Kids Under 6 \n(% samples >5µg/dL)") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(),
    panel.grid.major = element_blank(),  # removes major grid lines
    panel.grid.minor = element_blank(),  # already in your code; keeps it
    axis.title.y = element_text(margin = margin(r = 10))
  )

# Plot 3
p3 <- ggplot(school_plot, aes(x = med_income, y = value, color = med_income)) +
  stat_poly_eq(
    aes(group = 1, label = paste("bold(", ..rr.label.., ")")),
    formula = y ~ x,
    parse = TRUE,
    label.x = 110000,
    label.y = 25,
    size = 3,
    color = "gray30",
    geom = "label",
    color = "black",        # text color
    fill = "white",         # background box color
    label.size = 0,
    label.padding = unit(0.15, "lines")  # tighter box
  ) +
  geom_text_repel(
    aes(label = county),
    size = 2.5,
    color = "black",
    box.padding = 0.3,
    nudge_y = 0.3,
    segment.size = 0 
  ) +
  geom_smooth(method = "lm", se = FALSE, color = scales::alpha("gray30", 0.6), linewidth = 0.5) +
  geom_point(size = 2) +
  scale_color_viridis_c(option = "C", direction = -1, name = "Median Income") +
  scale_x_continuous(
    labels = scales::dollar_format(scale = 0.0001, suffix = "0K"),
    breaks = seq(60000, 130000, 10000)
  ) +
  labs(x = "Household Median Income", y = "Lead Contamination of School \nDrinking Water (% samples > 5 ppb)") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(),
    panel.grid.major = element_blank(),  # removes major grid lines
    panel.grid.minor = element_blank(),  # already in your code; keeps it
    axis.title.y = element_text(margin = margin(r = 10))
  )

# Combine the 3 plots vertically
final_plot <- p1 / p2 / p3 +
  plot_annotation(title = "          Lead Hazard Metrics vs. Household Median Income by County")
```

###**Plot number of flights versus % of fixtures over 5ppb and label high-risk schools (trisha)**
```{r}

fixture_data <- all_lead_datasets
  select(school_name, pct_above_5, county, school_district)%>%
  filter(county=="King")

#filtering out flights from only small airports
flight_data <- flight_data%>%
  filter(county=="King")%>%
  select(school_name, school_district, school_county, total_leaded_flights)


combined <- fixture_data %>%
  inner_join(flight_data, by="school_name")%>%
  mutate(label_override = case_when(
    school_name == "Rainier View Elementary School" ~ "Rainier View Elem",
    school_name == "Catharine Blaine K-8 School" ~ "Catharine Blaine K-8",
    school_name == "Maple Elementary School" ~ "Maple Elem",
    school_name == "Campbell Hill Elementary School" ~ "Campbell Hill Elem",
    school_name == "Showalter Middle School" ~ "Showalter Middle",
    school_name == "Foster Senior High School" ~ "Foster Senior High",
    school_name == "Raisbeck Aviation High School" ~ "Raisbeck Aviation High",
    TRUE ~ NA_character_
  ))

ggplot(data = combined, aes(x = total_leaded_flights, y = pct_above_5)) +
  geom_jitter(size = 3, width = 2000, height = 1, aes(color = school_district)) +
  scale_color_manual(
    values = c(
      "Auburn" = "#1f77b4",
      "Bellevue" = "#ff7f0e",
      "Enumclaw" = "#2ca02c",
      "Federal Way" = "#d62728",
      "Highline" = "#f781bf",
      "Kent" = "#8c564b",
      "Lake Washington" = "#FFF9C4",
      "Renton" = "#7f7f7f",
      "Seattle" = "#bcbd22",
      "Skykomish" = "#17becf",
      "Snoqualmie Valley" = "#4daf4a",
      "Tukwila" = "#D8BFD8"),
    name = "District"
  ) +
  geom_text_repel(
    data = subset(combined, relevant_flights >= 10000),
    aes(label = label_override),
    size = 2.5,
    color = "black",  
    nudge_y = 5,            
    vjust = 1,              
    direction = "y",        
    show.legend = FALSE, 
    segment.color = NA
  )+ 
  labs(
    title = "Lead Exposure in Schools Through Drinking Water and Air", 
    x = "\nLead-Fueled Aircrafts Flying Over School \n(# Flights)", 
    y = "Lead Contamination of School Drinking Water \n(% samples > 5 ppb)\n"
  ) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = "K"))+
  theme_minimal() + 
  theme(
    axis.line = element_line(color = "black"),
  ) 
```

###**Map certain selected schools in WA state and manually map relevant airports (trisha)**
```{r}
# Read in school point data
school_data <- all_lead_datasets %>%
  select(school_name, x, y)%>%
  filter(school_name %in% c(
    "Rainier View Elementary School",
    "Catharine Blaine K-8 School",
    "Maple Elementary School",
    "Campbell Hill Elementary School",
    "Showalter Middle School",
    "Foster Senior High School",
    "Raisbeck Aviation High School"
  ))

# Convert to sf object
school_points <- st_as_sf(school_data, coords = c("Longitude", "Latitude"), crs = 4326)

# Optionally, get WA outline from ESD boundaries (or use any WA polygon shapefile)
WSD_boundaries <- read_sf("C:/Users/trish/OneDrive/Desktop/SWELL/Washington_School_District_Boundaries (1).geojson")
wa_outline <- st_make_valid(WSD_boundaries) %>% 
  st_union()

airports <- data.frame(
  name = c("Renton Municipal Airport", "Boeing Field"),
  lat = c(47.48573197727368, 47.537063621341694),
  long = c(-122.21664424732815, -122.30393060499654)
)

# Convert to sf object
airport_sf <- st_as_sf(airports, coords = c("long", "lat"), crs = 4326)


# Plot
ggplot() +
  geom_sf(data = wa_outline, fill = "white", color = "black", lwd = 0.3) +  # blank map
  
  geom_sf(data = school_points, color = "green", size = 2, alpha = 0.7) +
  geom_text(data = airports, aes(x = long, y = lat, label = "✈️"), size = 6, family = "sans") +
  theme_void() + 
  theme(legend.position = "none")+
  labs(title = "School Locations in Washington",
       subtitle = "Plotted on a blank WA state map",
       x = NULL, y = NULL)



#to plot JUST king county (run above code too, just not the ggplot)


king_wsd <- sd_map %>%
  filter(County == "King") #or any county you want to filter out, assuming you have data for it

# Fix any geometry issues and merge polygons
king_outline <- king_wsd %>%
  st_make_valid() %>%
  st_union()


# Plot
ggplot() +
  geom_sf(data = king_outline, fill = "white", color = "black", size = 0.4) +
  geom_sf(data = school_points, color = "red", size = 2) +
  geom_sf(data = airport_sf, aes(geometry = geometry), inherit.aes = FALSE) +
  geom_text(data = airports, aes(x = long, y = lat, label = "✈️"), size = 6, family = "sans") +
  coord_sf(crs = st_crs(king_outline)) +
  theme_void() +
  labs(title = "Selected Schools & Airports in King County, WA")


